M.Sc. Proposal: \\ \bigskip Computational Electrokinetics

### 2011-12-02 notes/summary/ch02.tex
Solver docs.

\section{Numerics}

@@

Salt flux are discretized on grid cells' boundaries 
(using upwind scheme $\mathcal{U}$ for numerical stability at large cell Peclet number):
\begin{eqnarray}
J_r &=& -\cD_r(C) + \alpha V_r \cdot \mathcal{U}^{\bV}_r (C) \\
J_\theta &=& -\frac{\cD_\theta(C)}{r} + \alpha V_\theta \cdot \mathcal{U}^{\bV}_\theta (C) \\
\nonumber \mathcal{U}^{\bV}_r(C)[i,j] &=& C\left[i-\frac{\sign(V_r)}{2}, j\right] \\
\nonumber \mathcal{U}^{\bV}_\theta(C)[i,j] &=& C\left[i, j-\frac{\sign(V_\theta)}{2}\right] 
\end{eqnarray}

@@

Force components are discretized on velocity staggered grid, where
linear interpolation is used for Coulomb force.
\begin{eqnarray}
F_r &=& -\cD_r(P) 
          + \cL(V_r) - \frac{2}{r^2} V_r 
		  - \frac{2}{r^2 \sin\theta} \cI_r(\cD_\theta (V_\theta \sin\theta))
          + \cD_r(\varPhi) \cdot \cI_r(\cL(\varPhi)) \\
F_\theta &=& -\frac{\cD_\theta(P)}{r} 
		  + \cL(V_\theta) - \frac{F_\theta}{r^2 \sin^2\theta} 
		  + \frac{2}{r^2} \cI_\theta(\cD_\theta(F_r))
		  + \frac{\cD_\theta(\varPhi)}{r} \cdot \cI_\theta(\cL(\varPhi)) \\
\nonumber
\cL(f) &=& \frac{1}{r^2}\cD_r\pars{\cD_r(f) r^2} + 
\frac{1}{r^2 \sin\theta} \cD_\theta\pars{\cD_\theta(f) \cdot \sin\theta}
\end{eqnarray}

@@

The boundary conditions at the ion-exchange boundary ($r=1$) are:
\begin{eqnarray}
\cI_r(\varPhi) &=& -\cI_r(\log C) \\
\cD_r(\varPhi) &=& \cD_r(\log C) \\
V_r &=& 0 \\
V_\theta &=& 4\log\pars{\frac{1 + \exp\left\{\cI_\theta(\zeta)/2\right\}}{2}} \cdot 
			\cD_\theta(\zeta) \\
 \nonumber \zeta &=& - \log \gamma - \cI_r(\varPhi)
\end{eqnarray}
The equations are discretized by using ``ghost'' points, outside the interior of the domain:
\begin{eqnarray}
 \nonumber \frac{\varPhi[0,j] + \varPhi[1,j]}{2} &=& 
	-\frac{\log C[0,j] + \log C[1,j]}{2} \\
 \nonumber \frac{\varPhi[1,j] - \varPhi[0,j]}{\Delta r} &=& 
	\frac{\log C[1,j] - \log C[0,j]}{\Delta r} 
\end{eqnarray}
By adding and subtracting the equations, we have:
\begin{eqnarray}
\varPhi[0,j] &=& -\log\brcs{C[1,j]} \\
C[0,j] &=& \exp\brcs{-\varPhi[1,j]} \\
 \nonumber \zeta[j] &=& \cV -\frac{\varPhi[0,j] + \varPhi[1,j]}{2} 
                     = -\log\gamma -\frac{\varPhi[0,j] + \varPhi[1,j]}{2}
\end{eqnarray}
The slipping conditions can be written by:
\begin{eqnarray}
V_r[0,j] &=& 0 \\
\frac{V_\theta[0,j] + V_\theta[1,j]}{2} &=& V_\theta[j] = 
4\log\pars{\frac{1 + \exp\left\{\frac{1}{2}\frac{\zeta[j] + \zeta[j+1]}{2}\right\}}{2}} 
\cdot \frac{\zeta[j+1] - \zeta[j]}{\bar{\theta}_{j+1} - \bar{\theta}_{j}} \\
 \nonumber V_\theta[0,j] &=& 2 V_\theta[j] - V_\theta[1,j]
\end{eqnarray}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Solver Design}
\subsubsection{Newton Method}
An operator $\cO$ is defined by its input vector $\bx$ and its output 
$\by = \cO(\bx)$ is defined on a grid $\cG$.

@@

Thus, in order to solve the equation $\cO(\bx) = \bzero$, a Newton step can 
be computed by:
\begin{eqnarray}
\bzero = \cO(\bx_n + \Delta \bx_n) &\approx& \cO(\bx_n) + \bnabla \cO(\bx_n)\cdot\Delta\bx_n \\
\bnabla \cO(\bx_n) \cdot \Delta \bx_n &=& -\cO(\bx_n) \\
\bx_{n+1} &=& \bx_{n} + \Delta \bx_n \rightarrow
\end{eqnarray}

@@

When the method converges $\bx_n \rightarrow \bx_\infty$, 
the convergence is quadratic:
\begin{eqnarray}
\|\bx_{n+1} - \bx_\infty\| \le c\|\bx_{n} - \bx_\infty\|^2
\end{eqnarray}
Thus, when the initial guess is close enough to the solution, the solver
will converge in very few steps.

@@

The equations of the system (as described above) can be written as non-linear 
operators acting on the extended variables:
\begin{eqnarray}
0 &=& \cE_I (\tilde\varPhi, \tilde C) \mbox{ -- conservation of charge} \\
0 &=& \cE_J (\tilde C, \tilde\bV; \alpha) \mbox{ -- conservation of salt} \\
0 &=& \cE_M (\tilde\bV) \mbox{ -- conservation of mass} \\
0 &=& \cE_{\boldsymbol{p}} (\tilde\bV, \tilde P, \tilde\varPhi) 
\mbox{ -- conservation of momentum}
\end{eqnarray}
The problem can be casted as solving $\cO(\bx) = \bzero$ for $\cO = \cE \circ \cB$,
where $\cB$ is the ``ghost'' point extension operator and $\cE$ is the conservation
equations operator.

@@

Due to symmetry considerations, the total force $\bF$ is 
aligned with $\ui$ and it vanishes iff $\ui \cdot \bF = 0$.
\begin{eqnarray}
F_\imath = \ui \cdot \bF &=& \oint_\mathcal{S} \ui \cdot \tT \cdot \bnhat  dA = 
\int_0^\pi f_\imath(\theta) \cdot 2\pi \sin\theta d\theta 
\\ \nonumber 
f_\imath &=& \pars{-P + 2\cD_r(V_r) + 
\frac{1}{2}\pars{\cD_r(\varPhi)}^2 - \frac{1}{2r^2}\pars{\cD_\theta(\varPhi)}^2}\cos\theta 
\\ \nonumber 
&& -\pars{\cD_r(V_\theta) - \frac{V_\theta}{r}
+ \frac{1}{r}\cD_r(\varPhi) \cD_{\theta}(\varPhi)}\sin\theta
\end{eqnarray}

The integral above is approximated by 1D numerical quadrature, to yield 
the total force $F_\imath$ as a function of steady-state drift velocity $\cU$.
In order to solve for steady state, Secant method is applied to find
the root of $F_\imath(\cU) = 0$, given the problem parameters $\alpha, \beta, \gamma$.

@@

System steady-state drift velocity $\cU(\beta)$ is computed as a function 
of the electric field magnitude for many values of $\beta$.
Since the solution $\bx_\infty$ is continous in $\beta$, the results from previous 
iteration are used as an initial solution $\bx_0$ for the problem of new $\beta$ 
-- helping Newton Method convergence.

@@

>>>>

\section{Numerical Solver}

@@

Salt flux are discretized on grid cells' boundaries 
(using upwind scheme $\mathcal{U}$ for numerical stability at large cell Peclet number):
\begin{eqnarray}
J_r &=& -\cD_r(C) + \alpha V_r \cdot \mathcal{U}^{\bV}_r (C) \\
J_\theta &=& -\frac{\cD_\theta(C)}{r} + \alpha V_\theta \cdot \mathcal{U}^{\bV}_\theta (C) \\
 \mathcal{U}^{\bV}_r(C)[i,j] &=& C\left[i-\frac{\sign(V_r)}{2}, j\right] \\
 \mathcal{U}^{\bV}_\theta(C)[i,j] &=& C\left[i, j-\frac{\sign(V_\theta)}{2}\right] 
\end{eqnarray}

@@

Force components are discretized on velocity staggered grid, where
linear interpolation is used for Coulomb force.
\begin{eqnarray}
F_r &=& -\cD_r(P) 
          + \cL(V_r) - \frac{2}{r^2} V_r 
		  - \frac{2}{r^2 \sin\theta} \cI_r(\cD_\theta (V_\theta \sin\theta))
          + \cD_r(\varPhi) \cdot \cI_r(\cL(\varPhi)) \\
F_\theta &=& -\frac{\cD_\theta(P)}{r} 
		  + \cL(V_\theta) - \frac{F_\theta}{r^2 \sin^2\theta} 
		  + \frac{2}{r^2} \cI_\theta(\cD_\theta(F_r))
		  + \frac{\cD_\theta(\varPhi)}{r} \cdot \cI_\theta(\cL(\varPhi)) \\
\cL(f) &=& \frac{1}{r^2}\cD_r\pars{\cD_r(f) r^2} + 
\frac{1}{r^2 \sin\theta} \cD_\theta\pars{\cD_\theta(f) \cdot \sin\theta}
\end{eqnarray}

@@

The boundary conditions at the ion-exchange boundary ($r=1$) are:
\begin{eqnarray}
\cI_r(\varPhi) &=& -\cI_r(\log C) \\
\cD_r(\varPhi) &=& \cD_r(\log C) \\
V_r &=& 0 \\
V_\theta &=& 4\log\pars{\frac{1 + \exp\left\{\cI_\theta(\zeta)/2\right\}}{2}} \cdot 
			\cD_\theta(\zeta) \\
  \zeta &=& - \log \gamma - \cI_r(\varPhi)
\end{eqnarray}
The equations are discretized by using ``ghost'' points, outside the interior of the domain:
\begin{eqnarray}
  \frac{\varPhi[0,j] + \varPhi[1,j]}{2} &=& 
	-\frac{\log C[0,j] + \log C[1,j]}{2} \\
  \frac{\varPhi[1,j] - \varPhi[0,j]}{\Delta r} &=& 
	\frac{\log C[1,j] - \log C[0,j]}{\Delta r} 
\end{eqnarray}
By adding and subtracting the equations, we have:
\begin{eqnarray}
\varPhi[0,j] &=& -\log\brcs{C[1,j]} \\
C[0,j] &=& \exp\brcs{-\varPhi[1,j]} \\
  \zeta[j] &=& \cV -\frac{\varPhi[0,j] + \varPhi[1,j]}{2} 
                     = -\log\gamma -\frac{\varPhi[0,j] + \varPhi[1,j]}{2}
\end{eqnarray}
The slipping conditions can be written by:
\begin{eqnarray}
V_r[0,j] &=& 0 \\
\frac{V_\theta[0,j] + V_\theta[1,j]}{2} &=& V_\theta[j] = 
4\log\pars{\frac{1 + \exp\left\{\frac{1}{2}\frac{\zeta[j] + \zeta[j+1]}{2}\right\}}{2}} 
\cdot \frac{\zeta[j+1] - \zeta[j]}{\bar{\theta}_{j+1} - \bar{\theta}_{j}} \\
  V_\theta[0,j] &=& 2 V_\theta[j] - V_\theta[1,j]
\end{eqnarray}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Solver Design}
\subsubsection{Newton Method}
An operator $\cO$ is defined by its input vector $\bx$ and its output 
$\by = \cO(\bx)$.

@@

Thus, in order to solve the equation $\cO(\bx) = \bzero$, a Newton step can 
be computed by:
\begin{eqnarray}
\bzero = \cO(\bx_n + \Delta \bx_n) &\approx& \cO(\bx_n) + \bnabla \cO(\bx_n)\cdot\Delta\bx_n \\
\bnabla \cO(\bx_n) \cdot \Delta \bx_n &=& -\cO(\bx_n) \\
\bx_{n+1} &=& \bx_{n} + \Delta \bx_n 
\end{eqnarray}

@@

When the method converges $\bx_n \rightarrow \bx_\infty$, 
the convergence is quadratic:
\begin{eqnarray}
\|\bx_{n+1} - \bx_\infty\| \le \|\bx_{n} - \bx_\infty\|^2
\end{eqnarray}
Thus, when the initial guess is close enough to the solution, the solver
will converge in very few steps.

@@

The concatenated equations of the system (as described above) can be written 
as a non-linear operator $\cE(\tilde\bx) = \bzero$ acting on the extended variables $\tilde\bx = \cB(\bx)$:
The problem can be casted as solving $\cO(\bx) = \bzero$ for $\cO = \cE \circ \cB$,
where $\cB$ is the ``ghost'' point extension operator (derived using boundary conditions) 
and $\cE$ is the conservation equations operator (derived by concatenation of the system
equations).

@@

Due to symmetry considerations, the total force $\bF$ is 
aligned with $\ui$ and it vanishes iff $\ui \cdot \bF = 0$.
\begin{eqnarray}
F_\imath = \ui \cdot \bF &=& \oint_\mathcal{S} \ui \cdot \tT \cdot \bnhat  dA = 
\int_0^\pi f_\imath(\theta) \cdot 2\pi \sin\theta d\theta 
\\  
f_\imath &=& \imath \cdot \tT \cdot \bnhat
\\  
f_\imath &=& \pars{-P + 2\cD_r(V_r) + 
\frac{1}{2}\pars{\cD_r(\varPhi)}^2 - \frac{1}{2r^2}\pars{\cD_\theta(\varPhi)}^2}\cos\theta 
\\  
&& -\pars{\cD_r(V_\theta) - \frac{V_\theta}{r}
+ \frac{1}{r}\cD_r(\varPhi) \cD_{\theta}(\varPhi)}\sin\theta \\
 \mathcal{S} &=& \{\br : \|\br\|_2 = 1\}
\end{eqnarray}

The integral above is approximated by 1D numerical quadrature, to yield 
the total force $F_\imath$ as a function of the drift velocity $\cU$.

Since the goal is to find steady-state solution, $F_\imath(\cU)$ is required to be zero --
and the appropriate $\cU$ is found by 1D root-finding algorithm.

@@

System steady-state drift velocity $\cU(\beta)$ is computed as a function 
of the electric field magnitude for several values of $\beta$.
Since the solution $\bx_\infty$ is continous in $\beta$ and $\cU$, 
the result of the current iteration is used as an initial solution $\bx_0$ for the problem 
of the next iteration -- helping the convergence of the iterative solver in use.

@@

Object-Oriented Design methodology is used to for solver implementation.
MATLAB programming language is chosen due to strong numerical capabilities
and high-level language features.

\subsubsection{Operator Interface}
The base \verb|Operator| interface is defined by 
supporting the computation of a residual/result vector \verb|op.res()|
and a gradient matrix \verb|op.grad()|.
This interface is implemented in various operators classes,
which are used to constuct $\cO = \cE \circ \cB$ and to apply Newton method
to the problem by computing the residual and the gradient automatically.
Each such operator may be defined using other operators, so its residual
and gradient are computed recursively.

\subsubsection{Unary operators}
\paragraph{Linear operator}
If $\cO$ is linear, it can be represented by a matrix $L$, such that:
\begin{eqnarray}
\cO(\bx) &=& L \bx \\
\bnabla\cO &=& L 
\end{eqnarray}
Note that finite difference $\cD$, interpolation $\cI$ and 
upwind selection $\mathcal{U}$ operators
can be implemented as linear sparse operators, by constructing 
an appropriate sparse matrix $L$, having $O(\dim \bx)$ non-zeroes)

\paragraph{Pointwise scalar function}
Let $f: \R \rightarrow \R$ be a differentiable 1D function, whose derivative is
denoted by $f': \R \rightarrow \R$. The operator $\cF$ can be defined to
represent pointwise application of $f$:
\begin{eqnarray}
\cF(\bx) &=& [f(x_1); \ldots; f(x_n)] \\
\bnabla\cF(\bx) &=& \diag(f'(x_1), \ldots, f'(x_n))
\end{eqnarray}
$f$ is defined and differentiated automatically using MATLAB symbolic toolbox.

\paragraph{Constant value}
A constant operator $\mathcal{C}$ is defined as having constant vector value $\bc$ 
(that does not depend on $\bx$):
\begin{eqnarray}
\mathcal{C}(\bx) &=& \bc \\
\bnabla\mathcal{C} &=& \bzero
\end{eqnarray}

\subsubsection{Binary operators}
Let $\cO_1$ and $\cO_2$ be two operators, so their pointwise addition, subtraction
and multiplication are defined as follows:
\begin{eqnarray}
(\cO_1 \pm \cO_2)(\bx) &=& \cO_1)(\bx) \pm \cO_2(\bx) \\
\bnabla (\cO_1 \pm \cO_2)(\bx)   &=& \bnabla\cO_1(\bx) \pm \bnabla\cO_2(\bx) \\
(\cO_1 \cdot \cO_2)(\bx) &=& \diag(\diag(\cO_1(\bx)) \cdot \diag(\cO_2(\bx))) \\
\bnabla (\cO_1 \cdot \cO_2)(\bx) &=& \diag(\cO_2(\bx)) \bnabla\cO_1(\bx) 
                                  + \diag(\cO_1(\bx)) \bnabla\cO_2(\bx)
\end{eqnarray}

\subsubsection{N-ary operators}
Given $N$ operators $\{\cO_i\}_{i=1}^N$, their concatenation $\cO$ is defined as:
\begin{eqnarray}
\cO(\bx) &=& [\cO_1(\bx); \ldots; \cO_N(\bx)] \\
(\bnabla\cO)(\bx) &=& [(\bnabla\cO_1)(\bx); \ldots; (\bnabla\cO_N)(\bx)]
\end{eqnarray}

\subsubsection{Iterative Newton solver}
Given an initial solution $\bx_0$, we apply the following algorithm 
until convergence ($\|\Delta \bx\| < \eps$):
\begin{enumerate}
\item Update the system: $\verb|op| = \verb|System|(\bx_{n})$
\item Compute the gradient: $\boldsymbol{A}_n = \verb|op.grad()|$
\item Compute the right-hand side: $\boldsymbol{b}_n = \verb|-op.res()|$
\item Solve sparse linear system: $\boldsymbol{A}_n \Delta \bx_n = \boldsymbol{b}_n$
\item Update solution: $\bx_{n+1} = \bx_{n} + \Delta \bx_{n}$
\end{enumerate}

\subsubsection{Total force integration -- $F_\imath$}
After the solver converges ($\bx_n \rightarrow \bx_\infty$), 
the ``ghost'' points values are computed by applying $\cB$ and
solving for zero force  $\brhat$ component on $r=1$ boundary

$F_\imath$ is computed using mid-point quadrature method:
\begin{eqnarray}
F_\imath &=& \sum_{i=1}^{n_\theta} f(\bar\theta_i) \cdot 
              2 \pi \sin\bar\theta_i \cdot (\theta_i - \theta_{i-1})
\end{eqnarray}

\subsubsection{Steady state drift velocity -- $\cU$}
In steady state, the force acting on the particle $F_\imath$ shoud vanish.
This yields an equation on $\cU$:
\begin{eqnarray}
F_\imath(\cU) &=& 0
\end{eqnarray}
This equation can be solved using 1D root finding algorithm, e.g. Secant method:
\begin{eqnarray}
\cU_{n+1} = \frac{F_\imath(\cU_{n})\cU_{n-1} - F_\imath(\cU_{n-1})\cU_{n}}
{F_\imath(\cU_{n}) - F_\imath(\cU_{n-1})} \rightarrow \cU_\infty
\end{eqnarray}
The result $\cU_\infty$ is the 
steady state particle drift velocity for the electric field $\beta$.
